A Linear Inside-Outside Algorithm for Correcting Sequencing Errors in Structured RNAs

### 2013-05-23 JCB/conclusion_JCB.tex
Initial commit JCB extended version

%!TEX root = main_RECOMB.tex
\section{Conclusion}
\label{sec:conclusion}

In this article we presented a new and efficient way of exploring the mutational landscape of an RNA under structural constraints,
and apply our techniques to identify and fix sequencing errors. In addition, we introduce a new
scoring scheme combining the nearest-neighbour energy model to new isostericity matrices in order to account for geometrical
discrepancies occurring during base pair replacements.
The algorithm runs in  $\Theta(n\cdot(|\Omega|+m^2))$ time and $\Theta(n\cdot(|\Omega|+m))$ memory, where $n$ is the length of the RNA,
$m$ the number of mutations and $\Omega$ the size of the multiple sequence alignment.

By combining  into \texttt{RNApyro} these two approaches,  the 
mutational landscape exploration and the pseudo energy model,
 we created a tool predicting the positions
 yielding point-wise sequencing error and correcting them.
We validated our model with the 5s rRNA,
as presented in Sec.~\ref{sec:results}.
We observed that the models
with larger weights on the
isostericity seems to hold a higher accuracy on the estimation of errors.
This indicates that an extractable signal is contained in the isostericity.
Importantly, the implementation is fast enough for practical applications.

>>>>

%!TEX root = main_RNAPyro_JCB.tex
\section{Conclusion}
\label{sec:conclusion}

\TODOJerome{Etoffer la conclusion}
In this article we presented a new and efficient way of exploring the mutational landscape of an RNA under structural constraints,
and apply our techniques to identify and fix sequencing errors. In addition, we introduce a new
scoring scheme combining the nearest-neighbour energy model to new isostericity matrices in order to account for geometrical
discrepancies occurring during base pair replacements.
The algorithm runs in  $\Theta(n\cdot(|\Omega|+m^2))$ time and $\Theta(n\cdot(|\Omega|+m))$ memory, where $n$ is the length of the RNA,
$m$ the number of mutations and $\Omega$ the size of the multiple sequence alignment.

By combining  into \texttt{RNApyro} these two approaches,  the 
mutational landscape exploration and the pseudo energy model,
 we created a tool predicting the positions
 yielding point-wise sequencing error and correcting them.
We validated our model with the 5s rRNA,
as presented in Sec.~\ref{sec:results}.
We observed that the models
with larger weights on the
isostericity seems to hold a higher accuracy on the estimation of errors.
This indicates that an exploitable signal is captured by the isostericity.
Importantly, the implementation is fast enough for practical applications.

### 2013-05-23 JCB/results_JCB.tex
Initial commit JCB extended version
Added a few 'todo' notes

%!TEX root = main_RECOMB.tex
\section{Results}
\label{sec:results}

@@

The time benchmarks were performed on a MacMini 2010, 2.3GHz dual-core Intel Core i5, 8GB of RAM.
Since typical use-cases of \RNApyro require
 efficiency and scalability, we present in Table~\ref{tab:time} typical runtimes required to compute the probabilities for  every nucleotide at every positions for a vast set of parameters. For those tests,
 both the sequences and the target secondary structure were generated at random.

@@

\subsection{Error Correction in 5s rRNA}

@@

To evaluate our method, we computed a ROC curve representing the performance of a classifier based on the mutational probabilities computed
by \RNApyro. More specifically, we fixed a threshold $\lambda \in [0,1]$, and predicted an error at position $i$ in sequence $\omega$ if and only if the
probability $P(i,x)$ of a nucleotide $x \in \{ A,C,G,U \}$ exceeds this threshold. To correct the errors we used the set of nucleotides having probability
greated than $\lambda$, that is  $C(i) = \{ x \; | \;  x \in \{ A,C,G,U \} \mbox{ and } P(i,x) > \lambda \mbox{ and }  n \neq \omega[i] \},$ where $\omega[i]$ is
the nucleotide at position $i$ in the input sequence. We note that, for lower thresholds, multiple nucleotides may be available in $C(i)$ to correct
the sequence. Here, we remind that our aim is to estimate the potential of error-correction of \RNApyro, and not to develop a full-fledged error-correction pipe-line, which  
shall be the subject of further studies. Finally, we progressively varied $\lambda$ between $0$ and $1$ to calculate the ROC curve and the area
under the curve (AUC). Our results are reported in Figure~\ref{fig:ROCall}.

>>>>

%!TEX root = main_RNAPyro_JCB.tex
\section{Results}
\label{sec:results}

@@

The time benchmarks were performed on a MacMini 2010, 2.3GHz dual-core Intel Core i5, 8GB of RAM.
Since typical use-cases of \RNApyro require
 efficiency and scalability, we present in Table~\ref{tab:time} typical runtimes required to compute the probabilities for  every nucleotide at every positions for a vast set of parameters. For those tests,
 both the sequences and the target secondary structure were generated at random.
\TODOVlad{Add more extensive time benchmark, as a plot}

@@

\subsection{Error Correction in 5s rRNAs}

@@

To evaluate our method, we computed a ROC curve representing the performance of a classifier based on the mutational probabilities computed
by \RNApyro. More specifically, we fixed a threshold $\lambda \in [0,1]$, and predicted an error at position $i$ in sequence $\omega$ if and only if the
probability $P(i,x)$ of a nucleotide $x$ exceeds this threshold. To correct the errors we used the set of nucleotides having probability
greater than $\lambda$, that is  
$$C(i) = \{ x \; | \;  x \in \{ \Ab,\Cb,\Gb,\Ub \}, P(i,x) > \lambda \mbox{ and }  n \neq \omega[i] \},$$
 where $\omega[i]$ is
the nucleotide at position $i$ in the input sequence. We note that, for lower thresholds, multiple nucleotides may be available in $C(i)$ to correct
the sequence. Here, we remind that our aim is to estimate the potential of error-correction of \RNApyro, and not to develop a full-fledged error-correction pipe-line, which  
shall be the subject of further studies. Finally, we progressively varied $\lambda$ between $0$ and $1$ to calculate the ROC curve and the area
under the curve (AUC). Our results are reported in Figure~\ref{fig:ROCall}.

### 2013-05-23 Recomb/introduction_RECOMB.tex

Ribonucleic acids (RNAs) are found in every living organisms and exhibit a broad range of functions, from catalyzing
chemical reactions as the RNase P or the group II introns, hybridizing  messenger RNA to regulate gene expression,
to ribosomal RNA (rRNA) synthesizing proteins. Those functions  require specific structures,  encoded in their nucleotide
sequence. Although the functions, and thus the structures, need to be preserved through various organisms, the sequences
can greatly differ from one organism to another. This sequence diversity coupled with the structural conservation is a fundamental
asset for evolutionary studies. To this end, algorithms to analyze the relationship between RNA mutants and structures are required.

>>>>

Ribonucleic acids (RNAs) are found in every living organism, and exhibit a broad range of functions, ranging from catalyzing
chemical reactions, as the RNase P or the group II introns, hybridizing  messenger RNA to regulate gene expressions,
to ribosomal RNA (rRNA) synthesizing proteins. Those functions  require specific structures,  encoded in their nucleotide
sequence. Although the functions, and thus the structures, need to be preserved through various organisms, the sequences
can greatly differ from one organism to another. This sequence diversity coupled with the structural conservation is a fundamental
asset for evolutionary studies. To this end, algorithms to analyze the relationship between RNA mutants and structures are required.
